\documentclass{beamer}
\usepackage{amsmath}
\usepackage{formal-grammar,proof,forest}
\usepackage{./prelude}
\usepackage[mathrm=sym]{unicode-math}

\setlength{\parskip}{\baselineskip}

\newcommand{\num}[1]{\texttt{num}[#1]}
\newcommand{\ctx}[1]{\langle #1 \rangle}
\newcommand{\set}[1]{\{ #1 \}}

\title{Template}
\author{Ethan Zhang, Srujan Roplekar}
\institute[]{}
\date{}

\begin{document}

\frame{\titlepage}

\section{Introduction}

\begin{frame}
	\frametitle{What are Semantics?}
\end{frame}

\begin{frame}
	\frametitle{Why Study Semantics?}
\end{frame}

\begin{frame}
	\tableofcontents
\end{frame}

\section{Arithmetic Expressions}

\begin{frame}
	\frametitle{Arithmetic Expressions: Syntax}

	\begin{center}
		\begin{grammar}
			\firstcase{e}{\num{n}}{Numeric constants}
			\otherform{\nonterm{e} \mathrel{\texttt{+}} \nonterm{e}}{Addition}
			\otherform{\nonterm{e} \mathrel{\texttt{*}} \nonterm{e}}{Multiplication}
		\end{grammar}
	\end{center}

	In English, an expression $e$ is:
	\vspace{-\parskip}
	\begin{itemize}
		\item a numeric constant \num{$n$} where $n$ is a number,
		\item a sum $e \mathrel{\texttt{+}} e$ where each $e$ is an expression, or
		\item a product $e \mathrel{\texttt{*}} e$ where each $e$ is an expression.
	\end{itemize}
\end{frame}

\begin{frame}
	\frametitle{Arithmetic Expressions: Example}

	\begin{columns}
		\begin{column}{0.45\textwidth}
			\centering
			\texttt{1 + ((2 + 3) * 4)}
		\end{column}
		\begin{column}{0.1\textwidth}
			\centering
			$\longmapsto$
		\end{column}
		\begin{column}{0.45\textwidth}
			\centering
			\begin{forest}
				[
					\texttt{+}
					[
						\num{1}
					]
					[
						\texttt{*}
						[
							\texttt{+}
							[
								\num{2}
							]
							[
								\num{3}
							]
						]
						[
							\num{4}
						]
					]
				]
			\end{forest}
		\end{column}
	\end{columns}
\end{frame}

\begin{frame}
	\frametitle{Aside: (Big-step) Operational Semantics}

	The goal of \emph{operational semantics} is to describe a program by \textbf{what it does}.

	But what does this mean?
	One idea is to think of programs as \emph{functions}:
	given some inputs, we get some output.

	Following this intuition, big-step semantics describes a program by its final result.
	For example, the program $\texttt{1 + ((2 + 3) * 4)}$ is described by the number $21$.
\end{frame}

\begin{frame}
	\frametitle{Aside: Inference Rules}

	As computer scientists, we want to describe a program's semantics by formal rules, so that we can both reason about them mathematically and implement them in code.

	Following the tradition of logic, semantics are described by \emph{inference rules}:
	\[
		\infer[(\textsc{Rule-Name})]
		{
			J
		}
		{
			J_1 & J_2 & \dots & J_n
		}
	\]
	In English, we say that ``$J$ holds, provided that $J_1$, $J_2$, $\dots$, $J_n$ all hold.''
\end{frame}

\begin{frame}
	\frametitle{Arithmetic Expressions: Semantics}

	\[
		\infer[(\textsc{Const})]
		{
			\num{n} \Downarrow \num{n}
		}
		{}
	\]
	\vspace{.5em}
	\[
		\infer[(\textsc{Add})]
		{
			e_1 + e_2 \Downarrow \num{n + m}
		}
		{
			e_1 \Downarrow \num{n} & e_2 \Downarrow \num{m}
		}
		\qquad
		\infer[(\textsc{Mul})]
		{
			e_1 * e_2 \Downarrow \num{n \cdot m}
		}
		{
			e_1 \Downarrow \num{n} & e_2 \Downarrow \num{m}
		}
	\]
\end{frame}

\section{Adding Variables}

\begin{frame}
	\frametitle{Adding Variables: Syntax}

	\begin{center}
		\begin{grammar}
			\firstcase{e}{\num{n}}{Numeric constants}
			\otherform{\textcolor{UIUCorange}{x}}{\textcolor{UIUCorange}{Variable reference}}
			\otherform{\nonterm{e} + \nonterm{e}}{Addition}
			\otherform{\nonterm{e} * \nonterm{e}}{Multiplication}
			\otherform{\textcolor{UIUCorange}{\texttt{let $x$ = $\nonterm{e}$ in $\nonterm{e}$}}}{\textcolor{UIUCorange}{Variable definition}}
		\end{grammar}
	\end{center}

	In English, an expression $e$ is:
	\vspace{-\parskip}
	\begin{itemize}
		\item a numeric constant \num{$n$} where $n$ is a number,
		\item \textcolor{UIUCorange}{a variable reference $x$ where $x$ is an identifier,}
		\item a sum $e \mathrel{\texttt{+}} e$ where each $e$ is an expression,
		\item a product $e \mathrel{\texttt{*}} e$ where each $e$ is an expression, or
		\item \textcolor{UIUCorange}{a ``let'' expression, where we define a variable $x$.}
	\end{itemize}
\end{frame}

\begin{frame}
	\frametitle{Adding Variables: Example}

	\begin{columns}
		\begin{column}{0.47\textwidth}
			\centering
			\begin{minipage}{.61\textwidth}
				\texttt{let x = 4} \\
				\texttt{\phantom{l}in (let y = 5} \\
				\texttt{\phantom{ in (l}in (x + y))}
			\end{minipage}
		\end{column}
		\begin{column}{0.06\textwidth}
			\centering
			$\longmapsto$
		\end{column}
		\begin{column}{0.47\textwidth}
			\centering
			\begin{forest}
				[
					\texttt{let}
					[ \texttt{x} ]
					[ \num{4} ]
					[
						\texttt{let}
						[ \texttt{y} ]
						[ \num{5} ]
						[
							\texttt{+}
							[ \texttt{x} ]
							[ \texttt{y} ]
						]
					]
				]
			\end{forest}
		\end{column}
	\end{columns}
\end{frame}

\begin{frame}
	\frametitle{Adding Variables: Context}

	Now that we have variables, the value of a program depends not only on the source code but also on the \emph{environment} in which it is executed.
	For example, the value of the program \texttt{x + 5} depends on the current value of $x$!

	Hence, we introduce an \emph{execution context} to keep track of the all inputs we need.
	Previously, the context only program's source code.
	Now, it also contains a mapping of variable names to values:
	\[
		\ctx{ \text{program}, \text{environment} }
	\]
\end{frame}

\begin{frame}
	\frametitle{Adding Variables: Semantics}
	% FIXME: orange might not be the best color for additions...
	\[
		\infer[(\textsc{Const})]
		{
			\ctx{ \num{n}, S } \Downarrow \num{n}
		}
		{}
		\qquad
		\textcolor{UIUCorange}{\infer[(\textsc{Var})]
			{
				\ctx{ x, S } \Downarrow \num{n}
			}
			{
				x \mapsto \num{n} \text{ in } S
			}}
	\]
	\vspace{.5em}
	\[
		\infer[(\textsc{Add})]
		{
			\ctx{ e_1 + e_2, S } \Downarrow \num{m + n}
		}
		{
			\ctx{ e_1, S } \Downarrow \num{m} & \ctx{ e_2, S } \Downarrow \num{n}
		}
	\]
	\vspace{.5em}
	\[
		\infer[(\textsc{Mul})]
		{
			\ctx{ e_1 * e_2, S } \Downarrow \num{m \cdot n}
		}
		{
			\ctx{ e_1, S } \Downarrow \num{m} & \ctx{ e_2, S } \Downarrow \num{n}
		}
	\]
	\vspace{.5em}
	\[
		\textcolor{UIUCorange}{\infer[(\textsc{Let})]
			{
				\ctx{ \texttt{let $x$ = $e_1$ in $e_2$}, S } \Downarrow v
			}
			{
				\ctx{ e_1, S } \Downarrow v_1 & \ctx{ e_2, S\set{ x \mapsto v_1 } } \Downarrow v
			}}
	\]
\end{frame}

\end{document}
